\large{\bf Large effects and the infinitesimal model }

### 2023-11-30 review-responses.tex

Thanks for the careful reading and useful comments.

\begin{point}{}
    Paragraph starting ``This suggests exploring whether stable distributions..'' \revref:
The questions as stated: e.g., whether stable distributions can ``stand in'' for a Gaussian distribution in the infinitesimal model or whether the results of Barton et al ``carry over'' when the sum of effect sizes follows a stable law are quite vague and allow for multiple interpretations. Ideally, in order to address whether Barton et al's results (do not) carry over, one would like to at least show that with a power law distribution of effect sizes, the ``Mendelian noise term'', i.e., the difference between the offspring trait value and the midparent value, is (not) independent of the midparent value in the limit of a very large number of loci. And indeed this is what the authors address
somewhat non-rigorously (and for the case of a Cauchy distribution of effect sizes) in section 5 and using simulations : see e.g., figures 4C and 4f, figures 6C and 6F. Unsurprisingly, these simulations show that with e.g., Cauchy distribution of effect sizes, the non-independence between the segregation/noise term and the midparent trait value does not hold. However, a lot of the mathematical analysis in the paper (sections 3 and 4) explores when the distribution of trait values in a population is (not) Gaussian, which is a completely different question.
\end{point}

\reply{
    We don't actually agree with this characterization --
    sections 3 and 4 explore the consequences of different distributions
    of the Mendelian segregation term (which derives from the sum of effects),
    not the population distribution.
    However, the assumptions and motivation were not nearly as clear as they could have been
    in the previous draft, and this is hopefully more clear now --
    see responses to points below.
}

@@

\begin{point}{\revref}
In the same paragraph: I don't think the claim that ``independence of offsprings' deviations and midparent values implies a Gaussian distribution'' (which is also stated in the abstract) is true? Surely, ``independence of offsprings' deviations and midparent values'' is just the basic statement of the infinitesimal model (say, in Barton et al 2017) and the infinitesimal model does not imply a Gaussian distribution of trait values? See also section 3.
\end{point}

\reply{
    We have modified the sentence to say
    ``implies a Gaussian for the offsprings' deviations'',
    to make it clear that this statement does not refer to the population distribution,
    but rather to the Mendelian sampling term \revref.
    See additional points below on this topic.
}

@@

As an alternative, would it make sense to look at what fraction of loci have effect size difference between alternative alleles to be $>2t$ without weighing by allele frequency?
More broadly, how do we interpret the frequency-weighted vs. unweighted distributions?
\end{point}

\reply{
    Good question; this was somewhat obscure in the previous draft.
    We've added a comment which hopefully explains why the frequency-weighted distribution
    is the correct thing to consider for this calculation \revref,
    and have added some more discussion about the point in general
    to the Discussion \llname{discussion_DFEs}.
}

\begin{point}{}
I was also wondering about the ``goodness of fit'' of observed distributions to power laws. I am not necessarily recommending a detailed quantitative analysis, but nevertheless it may be useful to comment on whether one has power to say reject alternative distributions (e.g., a mixture of exponentials) that are often considered. If not, then perhaps the exact exponents inferred are unimportant, and what the data points to more broadly, is a broad distribution of effect sizes.
\end{point}

\reply{
    This is a good question, and we had in fact done some exploration of this question
    but not included it in the paper.
    We've still not done a detailed analysis,
    but now provide some discussion and numerical exploration of our precision here:
    see Figure~\ref{fig:power} and \revref.
}

@@

\begin{point}{}
    Introduction: Also worth citing Fisher, Bulmer etc. (the original references) when introducing the infinitesimal model?
\end{point}

\reply{
}

\begin{point}{\revref}
    ``Perhaps the distribution of effect sizes within each gene is Normal'': I am confused by this. Do you mean that the distribution of effect sizes across all genes (with roughly the same ``proximity'' to the trait) is normal. If yes, then the above phrasing is a bit misleading. If what is meant that is really that the effect sizes of different alleles within a gene are normally distributed, then I am not sure if this is entirely plausible: see various papers by Turelli (e.g., ``Heritable Genetic Variation via Mutation-Selection Balance: Lerch's Zeta Meets the Abdominal Bristle.''). Presumably, a normal distribution of effect sizes for a sufficiently large genomic region is a good approximation but it is unclear what is large enough: smaller or larger than a typical gene...?
\end{point}

\reply{
    We've rephrased this, and replaced ``gene'' with ``region'' --
    really, we just want to motivate a mixture model,
    as the "scale mixture of Gaussians" is a common way to get fat-tailed distributions,
    and eminently biologically plausible (we think).
}

@@

\begin{point}{\revref}
    Sentence starting ``In fact, many traits claimed...'': This phrasing here is rather cryptic and I do not understand what is being said. Maybe rephrase? Also a reference would help...
\end{point}

\reply{
}

\begin{point}{}
    Figure \ref{fig:exponent_hist}: Is there also a correlation between the number of cases and the number of SNPs? In other words, to what extent are (b) and (c) independent?
\end{point}

\reply{
    Yes; we've added this to Figure~\ref{fig:exponent_hist}.
}

@@

\begin{point}{}
    Last inequality on Page 13: should the subscript be $j$ instead of $i$?
\end{point}

\reply{
}

\begin{point}{}
    Simulations: Why not choose the same effect size distributions in the neutral case and the case with stabilizing selection? Alternatively, if different effect size distributions are used, it may be good to plot the distribution of (appropriately) scaled trait values. Right now the figures give the impression that more variation is maintained under stabilizing selection..
\end{point}

\reply{
    These are good suggestions, however, we have encountered difficulties in implementing this,
    so we've settled for a note alerting the reader not to read too much into the absolute scale
    (see Figure~\ref{fig:sel_trait_distrns} caption).
}

@@

\begin{point}{}
    Figure \ref{fig:seg_noise} caption: I find the phrasing ``(a,b) chosen to span the 5\% of midparents centered on the 10\% quantile of midparent value'' a bit confusing. Does this mean a and b are respectively the 7.5\% and 12.5\% quantile of the distribution of midparent trait values?
\end{point}

\reply{
    That's right; we've rephrased this confusing explanation.
}

### 2023-11-30 writeup.tex
more responses

As suggested by its name,
under the infinitesimal model the effects of any given genetic variant on the trait
is small (indeed, vanishing),
and selection-driven trait shifts are mediated by small changes in frequencies of many alleles of small effect.
However, empirical understanding of the genetic basis of trait variation in real-world species
often finds a significant portion of heritable variation explained by variants at only a few loci.
There are many possible explanations for this observation,
including selection or publication bias,
filters for larger-effect variants due to spatially varying selection
\citep{barton1987establishment,yeaman2011establishment,yeaman2011genetic},
and/or grouping together of concordant alleles by linkage or inversions
\citep{kirkpatrick2006chromosome}.
However,
it seems clear that in at least those cases where we understand the underlying molecular mechanisms,
mutations of large effect can be important in practice.
% \citep[e.g.,][]{cyclodiene-resistance,hoekstra2006genetics}.
The consensus in the field seems to be that traits exist on a spectrum between monogenic and polygenic,
and that many -- if not most -- lie in the intermediate region,
where heritable trait variation is due both to segregating large- and small-effect variants.
Indeed, some sensible verbal models of the biological mechanisms by which a genetic change
comes to have an effect on an organism's trait
describe a hierarchy or range of ``proximities'' to the trait,
by which changes to more ``nearby'' genes have the possibility of causing larger changes in the trait
\citep{kopp2012transcriptional,boyle2017expanded}.
This suggests a mixture model for the effect size distribution:
perhaps the distribution of effect sizes within each gene is Normal, \revpoint{1}{13}
but with a standard deviation that varies with ``proximity'' to the trait;
such mixtures can easily have large proportions of extreme values.

The Gaussian distribution appears thanks to ``the'' Central Limit Theorem,
i.e., the universality of adding up a great many very small and independent things.
However, there is a wider class of Central Limit Theorems,
which describe the distributions of sums of a great many independent and exchangeable things:
the \emph{stable laws} \citep{gnedenko1968limit}.
Roughly speaking, these say that if we add together $n$ independent copies of a random variable $X$
for which $\P\{ X > t \}$ is proportional to $t^{-\alpha}$ for some $0 < \alpha < 2$,
then -- almost regardless of the details -- the result, scaled by $n^{1/\alpha}$,
is well-approximated by a universal form, an $\alpha$-stable distribution
(and the approximation becomes exact, as $n \to \infty$).
Oth the other hand, if $\alpha \ge 2$ then the distribution has finite variance,
and so the sum of $n$, divided by $\sqrt{n}$, converges to a Gaussian distribution.
In such ensembles, single large entries can be important:
concretely, the largest of the $n$ copies will be of order $n^{1/\alpha}$,
so that e.g., for $\alpha=1$ a positive proportion of the sum of $n$ entries will be contributed by a single one.
On the other hand, the largest of $n$ Gaussians is of order $\sqrt(\log n)$. \revpoint{2}{1}

This suggests exploring whether stable distributions might reasonably stand in for the Gaussian
in the infinitesimal model. \revpoint{1}{1} \revpoint{1}{2}
In this paper, we make preliminary explorations in that direction:
Do the results of \citet{barton2017infinitesimal}
carry over when the distribution of effect sizes falls in the domain of attraction of a stable law?
The immediate answer to this will be ``no'' --
in fact, we will see that independence of offsprings' deviations and midparent values \emph{implies} a Gaussian distribution in some sense.
This independence is a key reason why the infinitesimal model is tractable,
and so a conclusion is that additive models with non-Gaussian effects may not be well-approximated
by ``trait-only'' models that do not explicitly represent underlying genetic variation.
Having thus shown that ``infinitesimal models with non-Gaussian noise'' is on weak footing as a model of reality,
we then go on to analyze exactly these models.
(As motivation,
we have not shown that such models are \emph{not} good approximations to reality,
and it is interesting and perhaps informative to understand how the noise distribution affects results.)

@@

\begin{figure}
    \begin{center}
    \includegraphics{snp_effects/results_10}
    \end{center}
    \caption{
        Estimated values of the tail exponent, $\alpha$,
        for 775 illness-related binary phenotypes:
        \textbf{(A)} distribution of values; and plotted against
        \textbf{(B)} number of SNPs, and
        \textbf{(C)} number of cases.
        \label{fig:exponent_hist}
    }
\end{figure}

@@

\begin{figure}
    \begin{center}
        \includegraphics{sims/neutral_seg_noise}
    \end{center}
    \caption{
        The distribution of Mendelian sampling terms
        with Gaussian \textbf{(A-C)} and Cauchy \textbf{(D-F)} effect size distributions,
        from the same simulations shown in Figure~\ref{fig:trait_distrns},
        for all individuals born during the last 1,000 time units of the simulation.
        \textbf{(A,D)} Mendelian noise (offspring trait minus midparent) plotted against midparent.
        \textbf{(B,E)} Histograms of the Mendelian sampling term.
        \textbf{(B,E)} Enrichment in quantiles of the Mendelian sampling term conditional on midparent value.
        To compute the latter,
        let $q_0, q_2, q_4, \ldots, q_{98}, q_{100}$ divide the distribution of the Mendelian sampling term
        into 50 bins with roughly equal numbers in each (i.e., the quantiles),
        let $p_k(a,b)$ denote the proportion of the offspring of parents with midparent value
        is in $[a, b)$ whose Mendelian sampling term is in $[q_{k-1}, q_k)$,
        and let $p_k$ be the proportion of all offspring with Mendelian noise in $[q_{k-1}, q_k)$
        (so, $p_k \approx 0.02$).
        Then, each line shows $p_k(a,b)/p_k$
        plotted against the midpoint of the corresponding quantile bin $(q_{k-1},q_k)$,
        for $(a,b)$ chosen to span the 5\% of midparents centered on:
        (red) the 10\% quantile,
        (black) the median, and
        (green) the 90\% quantile of midparent value.
        \label{fig:seg_noise}
    }
\end{figure}

@@

\begin{figure}
    \begin{center}
        \includegraphics{sims/selected_trait_traces}
    \end{center}
    \caption{
        As in Figure~\ref{fig:trait_distrns},
        except under a model of stabilizing selection (see text for details).
        \label{fig:sel_trait_distrns}
    }
\end{figure}

@@

>>>>

As suggested by its name,
under the infinitesimal model the effects of any given genetic variant on the trait
is small (indeed, vanishing),
and selection-driven trait shifts are mediated by small changes in frequencies of many alleles of small effect.
However, empirical understanding of the genetic basis of trait variation in real-world species
often finds a significant portion of heritable variation explained by variants at only a few loci.
There are many possible explanations for this observation,
including selection or publication bias,
filters for larger-effect variants due to spatially varying selection
\citep{barton1987establishment,yeaman2011establishment,yeaman2011genetic},
and/or grouping together of concordant alleles by linkage or inversions
\citep{kirkpatrick2006chromosome}.
However,
it seems clear that in at least those cases where we understand the underlying molecular mechanisms,
mutations of large effect can be important in practice.
% \citep[e.g.,][]{cyclodiene-resistance,hoekstra2006genetics}.
The consensus in the field seems to be that traits exist on a spectrum between monogenic and polygenic,
and that many -- if not most -- lie in the intermediate region,
where heritable trait variation is due both to segregating large- and small-effect variants.
Indeed, some sensible verbal models of the biological mechanisms by which a genetic change
comes to have an effect on an organism's trait
describe a hierarchy or range of ``proximities'' to the trait,
by which changes to more ``nearby'' genes have the possibility of causing larger changes in the trait
\citep{kopp2012transcriptional,boyle2017expanded}.
This suggests a mixture model for the effect size distribution:
perhaps a trait is affected by mutations in a number of genomic regions,
and the distribution of effect sizes within each region is Normal, \revpoint{1}{13}
but with a standard deviation that varies with ``proximity'' of that region to the trait
(e.g., regions that are further ``upstream'' have higher standard deviations).
Such mixtures can easily have large proportions of extreme values.

The Gaussian distribution appears thanks to ``the'' Central Limit Theorem,
i.e., the universality of adding up a great many very small and independent things.
However, there is a wider class of Central Limit Theorems,
which describe the distributions of sums of a great many independent and exchangeable things:
the \emph{stable laws} \citep{gnedenko1968limit}.
Roughly speaking, these say that if we add together $n$ independent copies of a random variable $X$
for which $\P\{ X > t \}$ is proportional to $t^{-\alpha}$ for some $0 < \alpha < 2$,
then -- almost regardless of the details -- the result, scaled by $n^{1/\alpha}$,
is well-approximated by a universal form, an $\alpha$-stable distribution
(and the approximation becomes exact, as $n \to \infty$).
Oth the other hand, if $\alpha \ge 2$ then the distribution has finite variance,
and so the sum of $n$, divided by $\sqrt{n}$, converges to a Gaussian distribution.
In such ensembles, single large entries can be important:
concretely, the largest of the $n$ copies will be of order $n^{1/\alpha}$,
so that e.g., for $\alpha=1$ a positive proportion of the sum of $n$ entries will be contributed by a single one.
On the other hand, the largest of $n$ Gaussians is of order $\sqrt{\log n}$. \revpoint{2}{1}

This suggests exploring whether stable distributions might reasonably stand in for the Gaussian
in the infinitesimal model. \revpoint{1}{1}
In this paper, we make preliminary explorations in that direction:
Do the results of \citet{barton2017infinitesimal}
carry over when the distribution of effect sizes falls in the domain of attraction of a stable law?
The immediate answer to this will be ``no'' --
in fact, we will see that independence of offsprings' deviations and midparent values \emph{implies} a Gaussian distribution for the offsprings' deviations, in some sense. \revpoint{1}{2}
This independence is a key reason why the infinitesimal model is tractable,
and so a conclusion is that additive models with non-Gaussian effects may not be well-approximated
by ``trait-only'' models that do not explicitly represent underlying genetic variation.
Having thus shown that ``infinitesimal models with non-Gaussian noise'' is on weak footing as a model of reality,
we then go on to analyze exactly these models.
(As motivation,
we have not shown that such models are \emph{not} good approximations to reality,
and it is interesting and perhaps informative to understand how the noise distribution affects results.)

@@

\begin{figure}
    \begin{center}
    \includegraphics{snp_effects/results_10}
    \end{center}
    \caption{
        Estimated values of the tail exponent, $\alpha$,
        for 775 illness-related binary phenotypes:
        \textbf{(A)} distribution of values; and plotted against
        \textbf{(B)} number of SNPs, and
        \textbf{(C)} number of cases.
        \textbf{(D)} shows the number of SNPs and number of cases by phenotype.
        \label{fig:exponent_hist}
    }
\end{figure}

A possible concern is that tail exponents are difficult to estimate accurately, \revpoint{1}{4}
but simulation-based estimates of accuracy (Supplemental Figure~\ref{fig:power})
suggest that our estimates are unbiased but only accurate to within roughly $\pm 0.4$,
so the distribution of Figure~\ref{fig:exponent_hist} should not be taken as
a direct estimate of the underlying distribution of $\alpha$ values,
but is robust evidence for exponents strictly between 1 and 2 (and probably around 1.5).

@@

\begin{figure}
    \begin{center}
        \includegraphics{sims/neutral_seg_noise}
    \end{center}
    \caption{
        The distribution of Mendelian sampling terms
        with Gaussian \textbf{(A-C)} and Cauchy \textbf{(D-F)} effect size distributions,
        from the same simulations shown in Figure~\ref{fig:trait_distrns},
        for all individuals born during the last 1,000 time units of the simulation.
        \textbf{(A,D)} Mendelian noise (offspring trait minus midparent) plotted against midparent.
        \textbf{(B,E)} Histograms of the Mendelian sampling term.
        \textbf{(B,E)} Enrichment in quantiles of the Mendelian sampling term conditional on midparent value.
        To compute the latter,
        let $q_0, q_2, q_4, \ldots, q_{98}, q_{100}$ divide the distribution of the Mendelian sampling term
        into 50 bins with roughly equal numbers in each (i.e., the quantiles),
        let $p_k(a,b)$ denote the proportion of the offspring of parents with midparent value
        is in $[a, b)$ whose Mendelian sampling term is in $[q_{k-1}, q_k)$,
        and let $p_k$ be the proportion of all offspring with Mendelian noise in $[q_{k-1}, q_k)$
        (so, $p_k \approx 0.02$).
        Then, each line shows $p_k(a,b)/p_k$
        plotted against the midpoint of the corresponding quantile bin $(q_{k-1},q_k)$,
        for $(a,b)$ chosen to span the three segments of 5\% of
        ``low'', ``medium'', and ``high'' midparents, i.e.:
        midparents between (red) the 7.5\% and 12.5\% quantiles,
        (black) the 47.5\% and 52.5\% quantiles, and
        (green) the 87.5\% and 92.5\% quantiles of midparent value.
        \label{fig:seg_noise}
    }
\end{figure}

@@

\begin{figure}
    \begin{center}
        \includegraphics{sims/selected_trait_traces}
    \end{center}
    \caption{
        As in Figure~\ref{fig:trait_distrns},
        except under a model of stabilizing selection
        (see text for details,
        and note that the scale of the underlying mutations do not match
        neutral simulations, so distributions are not directly comparable
        to Figure~\ref{fig:trait_distrns}).
        \label{fig:sel_trait_distrns}
    }
\end{figure}

@@

\begin{figure}
    \begin{center}
        \includegraphics{snp_effects/power_demo}
    \end{center}
    \caption{
        Distributions of estimated tail exponents using the same estimator
        as in Figure~\ref{fig:exponent_hist},
        for 1000 independent draws from
        \textbf{(left)} the Cauchy distribution (true $\alpha=1$),
        \textbf{(center)} the Student's $t$ distribution with 1.5 degrees of freedom (true $\alpha=1.5$), and
        \textbf{(right)} the Gaussian distribution (true $\alpha$ is undefined,
        as the tail decays super-polynomially).
        Note that the typical number of SNPs per trait (shown in Figure~\ref{fig:exponent_hist})
        is around 1000.
        \label{fig:power}
    }
\end{figure}
